---
title: 'Multivariate Time Series Analysis'
author: 'Dr. Yannick Hoga'
author2: 'Thilo Reinschlüssel'
subtitle: 'Solution Exercise Sheet 8'
semester: "Winter Term 2019/2020"
output:
  pdf_document:
    keep_tex: yes
    template: ../template.tex
    fig_caption: yes
    citation_package: biblatex
    number_sections: true
toc: true
lot: true
lof: true
graphics: true
linkcolor: black
urlcolor: black
citecolor: black
colorlinks: true
font: Times New Roman
fontsize: 12pt
geometry: lmargin = 2cm, rmargin = 2cm, tmargin = 2cm, bmargin = 2.5cm
classoption: a4paper
---

```{r , include=FALSE}
Sys.setlocale(locale = "English_United States.1252") ## English US Windows
knitr::opts_chunk$set(echo = TRUE)

source(here::here("packages/packages.R"))
```

# Exercise 1: Forecast Intervals and Distributional Assumptions

\begin{itemize}
  \item[a)] Which key assumption about the innovations $a_t$ is made in the lecture to derive the distribution of the forecast errors $e_T (h)$? 
\end{itemize}

_Solution:_

\begin{align*}
  a_t & \overset{\text{i.i.d.}}{\sim} N (0, \Sigma_a)
\end{align*}

\begin{itemize}
  \item i.i.d $\Rightarrow$ no autocorrelation
  \item $N \ \Rightarrow$ normal distributed 
  \item $\Sigma_a \ \Rightarrow$ heteroskedasticity
\end{itemize}

\begin{itemize}
  \item[b)] Assume we knew all parameters / coefficients and let $\Sigma_a$ be the identity matrix $I_{3 \times 3}$. Based on the assumption from a), derive the distribution of $e_T (1)$  for any stationary $\VAR(p)$.  
\end{itemize}

_Solution:_

\begin{align*}
  e_T (1) & = z_{T + 1} - z_T (1) \\
  & = a_{T + 1}\\
  \\
  & \text{holds for any } \VAR(p) \ \text{since}\\
  z_T (1) & = \mathbb{E} \left(z_{T + 1} | z_T, \ldots , z_0 \right)\\
  \\
  \Rightarrow e_t (1) & = a_{T + 1} \sim N \left(0, I_{3 \times 3} \right)
\end{align*}

\begin{itemize}
  \item[c)] Derive the confidence ellipsoid associated to b) for $\alpha = 5 \%$. What is the fraction forecast errors that lie inside the ellposid? 
\end{itemize}

_Solution:_

llipsoid: 

$$\left\{ z \in \mathbb{R}^{3}: \left( z_T (1) - z \right)^{'} \Sigma_e^{-1} (1) \left( z_T (1) - z \right) \leq \chi_{3, 1- \alpha}^{2}\right\}$$

By defining $z_T (1) - z =: \epsilon$ and using that $\Sigma_e (1) = \Sigma_a = I_{3 \times 3}$ the ellipsoid is: $\left\{ \epsilon \in \mathbb{R}^{3}: \epsilon^{'} \ \epsilon \leq \chi_{3, 1- \alpha}^{2}\right\}$. For $\alpha = 5 \%$, $95 \%$ of the observed forecast errors are expected to fall inside the confidence ellipsoid.

\begin{itemize}
  \item[d)] Run a simulation in `R `: Draw the forecast error $e_T (1)$ defined in a) and b) with $K = 3$. Check if it is located inside or outside the confidence ellipsoid derived in c). Use $N = 10 000$ repetitions in total and conclude whether the confidence ellipsoid is appropriate.  
\end{itemize}

_Solution:_

Just check if $e_T  (1)^{'} e_T (1)  \leq \chi_{3, 0.95}^{2}$ and compute $\displaystyle \dfrac{1}{N} \sum_{i = 1}^{N} \mathbbm{1} \left( e_T  (1)^{'} e_T (1)  \leq \chi_{3, 0.95}^{2}\right)$. 

```{r d}
N <- 10000 # number of repetitions
K <- 3 # dimension of VAR
 # drawing a_t from iid N(0,I)
gauss <- mvrnorm(n = N, mu = c(0,0,0), Sigma = diag(K))
# computing e'e for all draws in one take equals diag(EE')
msfe_gauss <- diag(gauss %*% t(gauss)) 
# this is only a one-sided test since we have squared each error!
limit <- qchisq(p = 0.95, df=3, lower.tail=TRUE) 
sum(msfe_gauss < limit) / N
```

\begin{itemize}
  \item[e)] Repeat the simulation from above but this time assume $a_t$ to be drawn from a uniform distribution. $\Sigma_a = I_{3 \times 3}$ remains. How reliable is the confidence ellipsoid in this case? 
  \textit{Hint: Set $\pm \dfrac{\sqrt{12}}{2}$ as lower / upper bound for unit variance.}
\end{itemize}

_Solution:_

```{r e}
# variance = 1 again, Kurtosis is < 3 for this one
unif <- matrix(data = runif(n = N * 3, min = -sqrt(12)/2, 
                            max = sqrt(12)/2), nrow = N, ncol = 3) 
msfe_unif <- rowSums(unif^2)
sum(msfe_unif < limit) / N
```

It is to conservative. 

\begin{itemize}
  \item[f)] Repeat the simulation drawing innovations from a $t$-distribution with $2$ degrees of freedom and conclude.
\end{itemize}

_Solution:_

```{r f}
# variance = 1 by default, kurtosis > 3 and this hurts a lot
t2 <- matrix(data = rt(n = N * 3, df = 2), nrow = N, ncol = 3)
msfe_t2 <- rowSums(t2^2)
sum(msfe_t2 < limit) / N
```

Too liberal, the ellipsoid is not appropriate. 



