---
title: 'Multivariate Time Series Analysis'
author: 'Dr. Yannick Hoga'
author2: 'Thilo Reinschlüssel'
subtitle: 'Solution Exercise Sheet 3'
semester: "Winter Term 2019/2020"
output:
  pdf_document:
    keep_tex: yes
    template: ../template.tex
    fig_caption: yes
    citation_package: biblatex
    number_sections: true
toc: true
lot: true
lof: true
graphics: true
linkcolor: black
urlcolor: black
citecolor: black
colorlinks: true
font: Times New Roman
fontsize: 12pt
geometry: lmargin = 2cm, rmargin = 2.5cm, tmargin = 2cm, bmargin = 2.5cm
classoption: a4paper
---

```{r , include=FALSE}
Sys.setlocale(locale = "English_United States.1252") ## English US Windows
knitr::opts_chunk$set(echo = TRUE)

source(here::here("packages/packages.R"))
```

# Exercise 1: VAR(1) Moments and Stationarity

Take the VAR(1) model $z_t = \phi_0 + \phi_1 z_{t-1} + a_t$ with the following parameterisation:

\begin{align*}
    \phi_0 = \begin{pmatrix} 1 \\ 0 \end{pmatrix}, \quad \phi_1 = \begin{pmatrix} 0.75 & 0 \\ -0.25 & 0.5 \end{pmatrix}, \quad \Sigma_a = \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}
\end{align*}

\begin{itemize}
    \item[a)] Compute the mean of the process.
\end{itemize}

_Solution:_

\begin{align*}
  \underset{\Rightarrow}{E(\cdot)} \quad \underbrace{ \mathbb{E} (z_t)}_{\mu} & = \mathbb{E} \left(  \underbrace{\phi_0}_{\phi_0} + \underbrace{\phi_1 z_{t -1}}_{\phi_1 \cdot \mu}+ \underbrace{a_t}_{0}\right)\\
  \left[ \text{Key assumption?} \;  \mathbb{E} (z_t)  = \mathbb{E} (z_{t-1}) \right]\\
  \Leftrightarrow \left( I - \phi_1 \right) \cdot \mu & = \phi_0\\
  \Leftrightarrow \mu & = \left( I - \phi_1 \right)^{-1} \cdot \phi_0\\
  \text{plugging in} \ \phi_0 \ \text{and} \ \phi_1:\\
  \mu & = \begin{pmatrix} 
  0.25 & 0 \\
  0.25 & 0.5
  \end{pmatrix}^{-1} 
  \begin{pmatrix} 
  1 \\
  0
  \end{pmatrix}\\
  & = 
  \begin{pmatrix} 
  -4 \\
  2
  \end{pmatrix}\\
\end{align*}

\begin{itemize}
    \item[b)] Show that the process is stationary.
\end{itemize}

_Solution:_ 

Eigenvalues of $\phi_1:$

\begin{align*}
  \left| \phi_1 - I \lambda \right| & \overset{!}{=} 0 \\
  & = \begin{vmatrix}
  0.75 - \lambda & 0 \\
  -0.25 & 0.5 - \lambda 
  \end{vmatrix}\\
  & = (0.75 - \lambda)(0.5 - \lambda) \overset{!}{=}  0 \\
  \Rightarrow \lambda_1 & = 0.75 \\
  \lambda_2 & = 0.5\\
\end{align*}

Both eigenvalues lie within the unit circle, therefore the process is stationary. 

\begin{itemize}
    \item[c)] Derive the Yule-Walker equations for the lags $l = \{0,1,2 \}$ and show that the solution for $\Gamma_0$ coincides with equation (2.3) on slide 2-15.
\end{itemize}

\begin{align*}
  \tilde{z}_t & := z_t - \mu \\
  \Rightarrow \tilde{z}_t & =  \phi_1 \tilde{z}_{t-1} + a_t \qquad & | \cdot \tilde{z}_{t-1}^{'}   \\
  \Leftrightarrow \tilde{z}_t \tilde{z}_t^{'} & = \phi_1 \tilde{z}_t \tilde{z}_t^{'} + a_t \tilde{z}_t^{'}\\
  \overset{\mathbb{E} (\cdot)}{\Rightarrow} \Gamma_l & = \phi_1 \cdot \Gamma_{l- 1} + \begin{cases} 
  l & = 0: \Sigma_a \\
  l & \neq 0: 0_k\\
  \end{cases}\\
  l & = 0: \Gamma_0 = \phi_1 \cdot \Gamma_{-1} + \Sigma_a\\
  l & = 1: \Gamma_1 = \phi_1 \cdot \Gamma_{0} + 0_K \\
  l & = 2: \Gamma_2 = \phi_1 \cdot \Gamma_{1} + 0_K \\
  \text{using} \ \Gamma_{-1} & = \Gamma_{1}^{'}\\ 
  \Gamma_0 & = \phi_1 \cdot \left( \phi_1 \Gamma_0 \right)^{'} + \Sigma_a\\
  \Leftrightarrow \Gamma_0 & = \phi_1 \Gamma_0^{'} \phi_1^{'}  + \Sigma_a\\
   \text{and} \ \Gamma_0^{'} & = \Gamma_0 \ \text{since}  \ \Gamma_0 \ \text{is symmetric !}
\end{align*}

\begin{itemize}
    \item[d)] Compute $\Gamma_0$ and $\Gamma_1$ by hand based on your results from c).
\end{itemize}

\begin{align*}
\\
  \Gamma_0 & = \Gamma_0^{'}\\
  \Rightarrow
  \begin{pmatrix} 
  \gamma_{11} & \gamma_{12}\\
  \gamma_{21} & \gamma_{22}
  \end{pmatrix} \ & =
  \begin{pmatrix} 
  \gamma_{11} & \gamma_{12}\\
  \gamma_{12} & \gamma_{22}
  \end{pmatrix}\\ 
\end{align*}


\begin{align*}
  & =
  \begin{pmatrix} 
  0.75 & 0\\
  -0.25 & 0.5
  \end{pmatrix} 
  \begin{pmatrix} 
  \gamma_{11} & \gamma_{12}\\
  \gamma_{12} & \gamma_{22}
  \end{pmatrix}
  \begin{pmatrix} 
  0.75 & -0.25\\
  0 & 0.5
  \end{pmatrix} +
  \begin{pmatrix} 
  1 & 0\\
  0 & 1
  \end{pmatrix}\\
  & = 
  \begin{pmatrix}
  (0.75)^2 \gamma_{11} & (-0.75 \cdot 0.25 \gamma_{11}) + 0.5 \cdot 0.75 \gamma_{12} \\
  (0.75 \cdot 0.25 \gamma_{11}) + 0.5 \cdot 0.75 \gamma_{12} & (-0.25)^2 \gamma_{11} - 0.25 \cdot 0.5 \gamma_{12} - 0.5 \cdot 0.25 \gamma_{12} + (0.5)^2 \gamma_{22}
  \end{pmatrix} +
  \begin{pmatrix}
  1 & 0\\
  0 & 1
  \end{pmatrix}
  \end{align*}

\begin{align*}
  \Rightarrow \gamma_{11} & = (0.75)^2 \gamma_{11} + 1\\
  \Leftrightarrow \gamma_{11} & = \dfrac{16}{7}\\
  \Rightarrow \gamma_{12} & = \dfrac{1}{4} \cdot \left( - \dfrac{3}{4} \right) \cdot \dfrac{16}{7} + \dfrac{2}{4} \cdot \dfrac{3}{4} \gamma_{12} \\
  \Leftrightarrow \gamma_{12} & = - \dfrac{24}{35}\\
  \Rightarrow \gamma_{22} & = \dfrac{1}{16} \gamma_{11} - 2 \cdot \dfrac{1}{2} \cdot \dfrac{1}{4} \gamma_{12} + \dfrac{1}{4} \gamma_{22} + 1 \\
  & =  \dfrac{1}{16} \cdot \dfrac{16}{7} - 2 \cdot \dfrac{1}{2} \cdot \dfrac{1}{4} \left(- \dfrac{24}{35} \right) + \dfrac{1}{4} \gamma_{22} + 1\\
  \gamma_{22} & = \dfrac{184}{105}\\
  \\
  \Gamma_1 & = \phi_1 \cdot \Gamma_0\\
  & = 
  \begin{pmatrix} 
  0.75 & 0 \\
  -0.25 & 0.5
  \end{pmatrix}
  \cdot 
  \begin{pmatrix} 
  \frac{16}{7} & \frac{-24}{35}\\
  \frac{-24}{35} & \frac{184}{105}
  \end{pmatrix}\\
  & = 
  \begin{pmatrix} 
  \frac{12}{7} & \frac{-18}{35}\\
  \frac{-32}{35} & \frac{22}{21}
  \end{pmatrix}
\end{align*}


# Exercise 2: Stationarity of VAR(p) Processes

Using the notation of Slide 2-27, prove that $|I_{kp} - \pmb{\Phi_1} z| = |I_k - \phi_1 z - \ldots - \phi_p z^p|$. Recall that $|A|$ denotes the determinant of the matrix $A$

_Hint: Derive_ $\pmb{\Phi}_1$ _and keep it mind that adding multiplies of columns/rows to other columns/rows does not affect the determinate! The plan is to end up with a special matrix._ 

_Solution:_ 

\begin{align*}
  z_t = & \phi_0  + \phi_1 z_{t-1} + \ldots + \phi_p z_{t - p} + a_t\\
  \text{as VAR (1):} & Z_t = \begin{pmatrix} \phi_0 \\ 
  0\\
  \vdots \\
  \end{pmatrix} + 
  \Phi_1 Z_{t - 1} +
  \begin{pmatrix} 
  a_t \\  
  0\\
  \vdots 
  \end{pmatrix}\\
  \text{with} & Z_t =   
  \begin{pmatrix} 
   z_t \\
   z_{t-1} \\
   \vdots \\
   z_{t-p+1}
  \end{pmatrix}
  \text{and} \ Z_{t-1} = 
  \begin{pmatrix} 
   z_t \\
   z_{t-1} \\
   \vdots \\
   z_{t-p}
  \end{pmatrix}\\
  \Phi_1 = &
  \begin{pmatrix}
  \phi_1 & \phi_2 & \ldots  & \ldots & \phi_p \\
  I_K & 0_K & \ldots & \ldots & 0_k\\
  0_K & I_K & 0_k & \ldots & 0_k\\
  \vdots & \ddots & \ddots &  & \vdots\\
  0_k & \ldots &  0_k & I_k & 0_k
  \end{pmatrix}
\end{align*}

$\Rightarrow$ Now one can simply apply the formula for a VAR(1) to check if the process is stationary.

\begin{align*}
\left| \Phi_1 - \lambda I \right| & \overset{!}{=} 0 \\
\Leftrightarrow (-1)^k \left| \Phi_1 - \lambda I \right| & \overset{!}{=} 0 \\
\Leftrightarrow \left| \lambda \left( I -  \Phi_1 \dfrac{1}{\lambda} \right)  \right| & \overset{!}{=} 0 \quad &  \text{since} \ \lambda \ \text{is a scalar} \ I = I_{kp} \\ 
\Leftrightarrow  \lambda^{kp} \left| I - \Phi_{1} \dfrac{1}{\lambda} \right|  & \overset{!}{=} 0 \quad & \text{and let} \dfrac{1}{\lambda} =: z \\
\Rightarrow \left| I - \Phi_1 z \right| & \overset{!}{=} 0 
\end{align*}

$\Rightarrow$ Stationarity if all  $|z_i| > 1 \; \, (|\lambda_i| < 1)$

\begin{align*}
  I_{kp} - \Phi_1 z & = 
  \begin{pmatrix}
    I_k - \phi_1 z & - \phi_2 z & - \phi_3 z & \ldots & - \phi_{p - 1} z & - \phi_{p} z \\
    - I_k z        & I_k        &  0_k       & \ldots & 0_k & 0_k \\
    0_k   & - I_k z        & I_k & \ldots & 0_k & \vdots\\
    \vdots & 0_k   & - I_k z        & \ldots & \vdots & \vdots \\
    0_k & 0_k & 0_k & \ldots &  - I_k z        & I_k
  \end{pmatrix}
\end{align*}

$\Rightarrow -I_k \cdot z  + I_k \cdot z = 0_k$

Since adding multiplies of columns to other columns does not affect the determinant: \

$\Rightarrow$ column 'i' $\cdot z$ + column 'i - 1' $\forall i \in \left\{ p, \ldots, 2\right\}$ yields a triangular matrix

\begin{align*}
  \Rightarrow
  \begin{pmatrix}
    I_k - \phi_1 z & - \phi_2 z^2 & \ldots & - \phi_{p} z & -\phi_1 & -\phi_2 z - \ldots -\phi_p z^{p-1}  \\
    & 0_k & & & I_k & &\\
    & 0_k & & & & I_k &\\
  \end{pmatrix}
\end{align*}

\begin{align*}
  \Rightarrow \left| I_{kp} - \Phi_1 z \right| & = \left( I_K - \phi_1 z - \ldots - \phi_p z^p \right) \cdot \prod_{i = 1}^{p-1} I_k & \text{since it is a triangular matrix}\\
  & = \left| I_k - \phi_1 z - \ldots - \phi_p z^p \right| 
\end{align*}

# Exercise 3: VAR(2) Moments and Stationarity

Consider the following VAR(2) model with i.i.d. innovations: 

\begin{align*}
    \phi_0 = \begin{pmatrix} 2 \\ 1 \end{pmatrix}, \quad \phi_1 = \begin{pmatrix} 0.5 & 0.1 \\ 0.4 & 0.5 \end{pmatrix}, \quad \phi_2 = \begin{pmatrix} 0 & 0 \\ 0.25 & 0 \end{pmatrix}, \quad \Sigma_a = \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}
\ \text{.}
\end{align*}


\begin{itemize}
    \item[a)] Show that the process is stationary.
\end{itemize}

_Solution:_ 

\begin{align*}
  \left| I_2 - \phi_1 z - \phi_2 z^2 \right| & \overset{!}{=} 0 \\
  \Rightarrow \begin{pmatrix} 1 & 0 \\ 
  0 & 1 \end{pmatrix} - \begin{pmatrix}  0.5 z & 0.1 z \\
  0.4 z & 0.5 z
  \end{pmatrix} - 
  \begin{pmatrix}  0 & 0 \\
  0.25 z & 0
  \end{pmatrix} & \overset{!}{=} 0\\
  \Leftrightarrow
  \left|
    \begin{pmatrix}  1- 0.5 z & -0.1 z \\
      -0.4 z - 0.25z^2 & 1- 0.5z
  \end{pmatrix}
  \right| & \overset{!}{=} 0\\
  = (1 - 0.5 z)^2 - (0.4z + 0.25z^2) \cdot 0.1z  & \overset{!}{=} 0 \\
  = 1 - z + 0.21 z^2 -0.025 z^3 & \overset{!}{=} 0
\end{align*}


```{r }
roots <- polyroot(c(1, -1, 0.21, 0.025))
roots # there are some imaginary parts attached to it
sum(abs(roots) < 1) # count how many roots lie inside the unit circle
```
Since all roots are outside the unit circle the process is stationary. Alternatively, we could also use the VAR(1) approach and compute the eigenvalues.

```{r }
phi_1 <- matrix(data = c(0.5, 0.4, 0.1, 0.5), nrow = 2)
phi_2 <- matrix(data = c(0, 0.25, 0, 0), nrow = 2)
I2x2 <- diag(2)
O2x2 <- matrix(data = rep(0, 4), nrow = 2)
Phi <- rbind( cbind(phi_1, phi_2), cbind(I2x2, O2x2) )
Phi
var1.eigen <- eigen(Phi)
sum(abs(var1.eigen$values) < 1) # How many eigenvalues lie inside the unit circle?
```

All `r sum(abs(var1.eigen$values) < 1)` eigenvalues are within the unit circle, so the process is stationary and we get the same result as before.

\begin{itemize}
    \item[b)] Determine the mean vector.
\end{itemize}

_Solution:_

\begin{align*}
  z_t & = \phi_0 + \phi_1 z_{t-1} + \phi_2 z_{t-2} +a_t\\
  \overset{E(\cdot)}{\Rightarrow} \ldots  \Rightarrow \mu & = \left( I - \phi_1 - \phi_2 \right)^{-1} \phi_0 \\  
  & = \begin{pmatrix}
  1 - 0.5 &  -0.1\\
  -0.4 - 0.25 & 1 - 0.5
  \end{pmatrix}^{-1}
   \begin{pmatrix}
  2\\
  1
  \end{pmatrix}
\end{align*}

```{r }
phi_0 <- c(2,1)
mu <- solve((I2x2 - phi_1 - phi_2)) %*% phi_0
fractions(mu)
```

\begin{itemize}
    \item[c)] Derive the Yule-Walker equations for the lags $l = \{0,1,2 \}$ for a general VAR(2) process.
\end{itemize}

_Solution:_

\begin{align*}
  \mathbb{E} \left( \tilde{z}_{t} \tilde{z}_{t-l}^{'} \right) & =  \phi_1 \mathbb{E} \left( \tilde{z}_{t} \tilde{z}_{t-l}^{'} \right) + \phi_2 \mathbb{E} \left( \tilde{z}_{t} \tilde{z}_{t-l}^{'} \right) + \mathbb{E} \left( a_{t} \tilde{z}_{t-l}^{'} \right)\\
  \\
  \Rightarrow l & = 0: \Gamma_0 = \phi_1 \Gamma_{-1} + \phi_2 \Gamma_{-2} + \Sigma_a \\
  l & = 1: \Gamma_1 = \phi_1 \Gamma_{0} + \phi_2 \Gamma_{-1} + 0_{2 \times 2}\\
  l & = 2: \Gamma_1 = \phi_1 \Gamma_{1} + \phi_2 \Gamma_{0} + 0_{2 \times 2}
\end{align*}

\begin{itemize}
    \item[d)] Suppose we only knew $\Gamma_0 , \Gamma_1$ and $\Gamma_2$ - how can we estimate $\phi_1$ and $\phi_2$ from it?
\end{itemize}

_Solution:_ 

\begin{align*}
  \overbrace{\begin{pmatrix} \Gamma_1 & \Gamma_2 \end{pmatrix}}^{\text{row vector}} & = \begin{pmatrix} \phi_1 & \phi_2 \end{pmatrix} \begin{pmatrix} \Gamma_0 & \Gamma_1\\
  \Gamma_1^{'} & \Gamma_0 \end{pmatrix}\\
  \Rightarrow \begin{pmatrix} \phi_1 & \phi_2 \end{pmatrix} & = \begin{pmatrix} \phi_1 & \phi_2 \end{pmatrix}^{-1} \begin{pmatrix} \Gamma_1 & \Gamma_2 \end{pmatrix} 
\end{align*}

\begin{itemize}
    \item[e)] Write the process as a VAR(1) and calculate the mean vector again.
\end{itemize}

_Solution:_

\begin{align*}
  Z_t = \begin{pmatrix}  z_{t} \\ z_{t-1} \end{pmatrix}\\
  \\
  \begin{pmatrix} z_t  \\ z_{t-1} \end{pmatrix} & = \begin{pmatrix} \phi_0  \\ 0_{2 \times 1 }\end{pmatrix} + \begin{pmatrix} \phi_1 & \phi_2  \\ I_{2 \times 2}  & 0_{2 \times 2} \end{pmatrix} \begin{pmatrix} z_{t-1}  \\ z_{t-2} \end{pmatrix} + \begin{pmatrix} a_t  \\ 0_{2 \times 1} \end{pmatrix} \\
  \overset{\mathbb{E} (\cdot)}{\Rightarrow} \begin{pmatrix} \mu  \\ \mu \end{pmatrix} & =\begin{pmatrix} \phi_0  \\ 0_{2 \times 1} \end{pmatrix} + \begin{pmatrix} \phi_1 & \phi_2 \\ I_{2 \times 2} & 0_{2 \times 2} \end{pmatrix} \begin{pmatrix} \mu  \\ \mu \end{pmatrix} + \begin{pmatrix} 0_{2 \times 2} \\ 0_{2 \times 2} \end{pmatrix}\\
  \begin{pmatrix} \mu  \\ \mu \end{pmatrix} & = \left[  \begin{pmatrix} I_{2 \times 2} & 0_{2 \times 2}\\ 0_{2 \times 2} & I_{2 \times 2} \end{pmatrix} - \begin{pmatrix} \phi_1 & \phi_2 \\ I_{2 \times 2} & 0_{2 \times 2} \end{pmatrix} \right]^{-1} \begin{pmatrix} \phi_0 \\ 0_{2 \times 2} \end{pmatrix} \\
  & = \begin{pmatrix} 
  1-0.5 & -0.1 & 0 & 0 \\
  -0.4 & 1-0.5 & -0.25 & 0 \\
  -1 & 0 & 1 & 0 \\
  0 & -1 & 0 & 1 
  \end{pmatrix}^{-1}
  \begin{pmatrix} 
  2 \\ 1 \\ 0 \\ 0 
  \end{pmatrix}
\end{align*}


```{r }
# Alternative solution to b) using the VAR(1) representation:
mu2 <- solve(diag(4) - Phi) %*% c(phi_0, rep(0,2))
fractions(mu2)
```


\begin{itemize}
    \item[f)] Compute $\Gamma_0$ based on the VAR(1) formulation.
    \item[] \textit{Hint: You can use R for the calculations.}
\end{itemize}
 
 _Solution:_
 
\begin{align*}
  \Gamma_0 & = \Phi \Gamma_0 \Phi + \Sigma_b \\
  \\
  \text{with} & \Phi = \begin{pmatrix} \phi_1 & \phi_2 \\
  I_{2 \times 2} & 0_{2 \times 2}
  \end{pmatrix}
  \end{align*}
  
\begin{align*}
  \text{and} \ \tilde{Z}_t \tilde{Z}_t & = \begin{pmatrix} \tilde{z}_t \\ \tilde{z}_{t -1} \end{pmatrix}   \begin{pmatrix} \tilde{z}_t & \tilde{z}_{t -1} \end{pmatrix} \\
  & =  \begin{pmatrix} \tilde{z}_t \tilde{z}_t^{'} & \tilde{z}_t \tilde{z}_{t -1}^{'} \\
   \tilde{z}_{t -1} \tilde{z}_t^{'} & \tilde{z}_{t -1} \tilde{z}_{t -1}^{'} \end{pmatrix} \\
   & = \underbrace{\begin{pmatrix} \Gamma_0 & \Gamma_1 \\ \Gamma_1^{'} & \Gamma_0 \end{pmatrix}}_{=: \Gamma_0^{*}}\\
   \\
   \text{gives:} & \left( I_{4 \times 4} - \Phi \otimes \Phi  \right) \ \vec(\Gamma_0^{*}) = \ \vec(\Sigma_b)\\
   \Leftrightarrow \vec(\Gamma_0^{*}) & = \left( I_{4 \times 4} - \Phi \otimes \Phi  \right)^{-1}  \ \vec(\Sigma_b)
\end{align*}

$\Rightarrow$ extract top left or bottom right matrix

```{r }
Sigma_a <- diag(2)
Sigma_b <- rbind( cbind(Sigma_a, O2x2), cbind(O2x2, O2x2) )
Gamma0ast.mat <- matrix(solve(diag(16) - Phi %x% Phi) %*% as.vector(Sigma_b)
                        , nrow = 4)

Gamma0ast.mat
```











