---
title: 'Multivariate Time Series Analysis'
author: 'Dr. Yannick Hoga'
author2: 'Thilo Reinschlüssel'
subtitle: 'Solution Exercise Sheet 4'
semester: "Winter Term 2019/2020"
output:
  pdf_document:
    keep_tex: yes
    template: ../template.tex
    fig_caption: yes
    citation_package: biblatex
    number_sections: true
toc: true
lot: true
lof: true
graphics: true
linkcolor: black
urlcolor: black
citecolor: black
colorlinks: true
font: Times New Roman
fontsize: 12pt
geometry: lmargin = 2cm, rmargin = 2cm, tmargin = 2cm, bmargin = 2.5cm
classoption: a4paper
---

```{r , include=FALSE}
Sys.setlocale(locale = "English_United States.1252") ## English US Windows
knitr::opts_chunk$set(echo = TRUE)

source(here::here("packages/packages.R"))
```

# Exercise 1: Information Criteria

Prove Corollary 4.5 from Slide 4-7.


_Solution:_

From Theorem 4.4:

\begin{align*}
  C(l) = \log \left(\hat{\Sigma_a} (l) \right) + \dfrac{l}{T} \cdot c_T
\end{align*}


\begin{itemize}
  \item[i)] $\lim \limits_{T \rightarrow \infty} c_T \longrightarrow \infty$ 
  \item[ii)] $\lim \limits_{T \rightarrow \infty} \dfrac{c_T}{T} \longrightarrow 0$
\end{itemize}

If i) and ii) hold, $C(l)$ chooses the optimal/correct model. 

\begin{itemize}
  \item AIC: $c_T  = 2K^2$
  \begin{align*}
    \lim \limits_{T \rightarrow \infty} c_T & = 2 \ K^2 \centernot\implies \infty
  \end{align*}
  \begin{itemize}
    \item[$\Rightarrow$] not consistent
  \end{itemize}
  \item BIC: $c_T = \log(T)  \cdot K^2$
    \begin{align*}
    \lim \limits_{T \rightarrow \infty} c_T & = \log(T) \cdot K^2   \implies \infty
    \lim \limits_{T \rightarrow \infty} \dfrac{c_T}{T} \dfrac{\log(T)}{T} K^2 \implies 0 
  \end{align*}
  \begin{itemize}
    \item[$\Rightarrow$] consistent
  \end{itemize}
  \item HQ: $c_T = 2 \ \log(\log(T)) \ K^2$
  \begin{align*}
    \lim \limits_{T \rightarrow \infty} c_T & = 2 \ \log(\log(T)) \ K^2 \implies \infty
    \lim \limits_{T \rightarrow \infty} \dfrac{c_T}{T} \dfrac{2 \ \log(\log(T)) \ K^2}{T} \implies 0 
  \end{align*}
  \begin{itemize}
    \item[$\Rightarrow$] consistent
  \end{itemize}
\end{itemize}


# Exercise 2: VAR(p): Data application
  
This exercise is concerned with finding an appropriate $\VAR(p)$ model for US macroeconomic data. You can find the dataset `us_macrodata.Rda` attached to this exercise sheet in the Moodle folder for this tutorial. Please use the `load` command to import the dataset from your directory into R.  There are 5 variables – CPI, Real GDP, the unemployment rate, general private investment and the debt-to-GDP ratio. All series have been sampled quarterly and were seasonally adjusted before downloaded from FRED.

```{r}
# loading data
load(file = here::here("exercise_MTSA/00_data/us_macrodata.Rda"))
# loading the MTS package
library(MTS)
```

\begin{itemize}
  \item[a] Plot all time series and judge which time series seem non-stationary. Proceed to compute growth rates of the non-stationary variables.
\end{itemize}

_Solution:_

```{r}
macmat <- data.matrix(us.macro_series)
plot.ts(macmat)
```

Every series except unemployment looks non-stationary. Regarding the debt-to-gdp ratio, this is surprising, but we better difference it as well.


```{r}
macdata <- cbind(diff(log(us.macro_series$cpi)), 
                 us.macro_series$unemprate[-(nrow(macmat))], 
                 diff(log(us.macro_series$rgdp)),
                 diff(log(us.macro_series$gp_investment)), 
                 diff(log(us.macro_series$debt_gdp)))

plot.ts(macdata)
```

Note that the last observation of 'unemp' was dropped for conformable length. Its last and not first due to the date information: measurements are always from the first day of a quarter.


\begin{itemize}
  \item[b] Perform a Ljung-Box test on the dataset. Does it look worthwhile to estimate a $\VAR(p)$
\end{itemize}

_Solution:_

```{r}
mq(x = macdata, lag = 20)
```

There is some correlation in the dataset. 


```{r}
mq(x = macdata[,-3], lag = 20)
```

Even without unemployment, there is some correlation in the dataset. 


\begin{itemize}
  \item[c] Determine the length of the time series. How many coefficients can be estimated and what does it mean for $K$ and $p$?
\end{itemize}

```{r}
data_dim <- dim(macdata) 

Tmax <- data_dim[1] # observations
K <- data_dim[2] # variables
(max.p <- (Tmax * K - K) / K^2 )
```

`r floor(max.p)` lags can be estimated in addition to the intercept. 





