---
title: "exercise_01"
author: "Jens_klenke"
date: "3 3 2020"
output: 
  pdf_document:
     includes:
        in_header: packages.sty
        keep_text: yes
---

# Exercise 1

_Prove properties 3, 4 and 5 from Proposition 1.2 (Slide 1-11). Are there any requirements regarding the matrix dimensions?_ 

\begin{enumerate}[i]
  \item One
  \item Two
  \item Three
\end{enumerate}

# Exercise 2

_Find the extrema of the following functions (using pen and paper). Determine whether these points constitute minima, maxima or saddle points:_ 



\begin{itemize}
    \item[a)] $f(x,y) = (x -2)^2 + (y -5)^2 + xy$
    \item[b)] $g(x,y) = (x -1)^3 - (4y + 1)^2$
\end{itemize}



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
