---
title: "Exercise Sheet 13 - Task 2"
output: pdf_document
---

## Case 1: Intercept-only (No Constant)
```{r}
set.seed(123)
a_t   <- d.z_t <- z_t <- rbind(rnorm(200), rnorm(200))       # errors to generate first-order VECM
alpha <- matrix(c(-0.1, 0.1), nrow = 2, ncol = 1)            # loading matrix
beta  <- matrix(c(1,-1), nrow = 2, ncol = 1)                 # cointegrating matrix
for(i in 2:200){
   d.z_t[, i] <- alpha %*% t(beta) %*% z_t[, i-1] + a_t[, i] # simulating \Delta x_t
   z_t[, i]   <- z_t[, i-1] + d.z_t[, i]                     # adding Delta x_t to x_t-1 to obtain x_t
}
```
#### Extracting the common trend and plotting
```{r}
# Extracting the common trend:
alpha.orth <- c(-1,alpha[1]/alpha[2])                        # following the formula on slide 7-38
t(alpha.orth) %*% alpha                                      # making sure it is orthogonal
CST <- t(alpha.orth) %*% z_t                                 # extracting the common stochastic trend (CST)

# Plotting all into one picture:
par(mfrow=c(3, 1))
par(mar=c(0,3.1,3.1,0.1))
plot(z_t[1, ], type="l", xaxt="n", ylab = "$z_t$", main="")
mtext("No Constant", side = 3, line=1)
lines(z_t[2, ], col="red")
par(mar=c(0,3.1,0,0.1))
plot(1:200, t(beta) %*% z_t, type="l", xlab = "", ylab = "$\\beta^{'} z_t$", xaxt="n",)
par(mar=c(3.1,3.1,0,0.1))
plot(1:200, CST, xlab = "", ylab = "$\\alpha_{\bot}^{'} z_t$", type = "l")
mtext(expression(italic(t)), side = 1, line=2)
```

## Case 2: Intercept-only (Restricted Constant)
```{r}
mu0 <- matrix(c(1, 5), nrow = 2, ncol = 1)
nu0 <- - alpha %*% t(beta) %*% mu0
print(cat("mu0:", mu0, "nu0:", nu0))
for(i in 2:200){
   d.z_t[, i] <- nu0 + alpha %*% t(beta) %*% z_t[, i-1] + a_t[, i]
   z_t[, i]   <- z_t[, i-1] + d.z_t[, i]
}
```
#### Extracting the common trend and plotting
```{r}
# Extracting the common trend:
CST <- t(alpha.orth) %*% z_t 

# Plotting all into one picture:
par(mfrow=c(3, 1))
par(mar=c(0,3.1,3.1,0.1))
plot(z_t[1, ], type="l", xaxt="n", ylab = "$z_t$", main="", ylim = c(min(z_t), max(z_t)))
mtext("Restricted Constant", side = 3, line=1)
lines(z_t[2, ], col="red")
par(mar=c(0,3.1,0,0.1))
plot(1:200, t(beta) %*% z_t, type="l", xlab = "", ylab = "$\\beta^{'} z_t$", xaxt="n",)
par(mar=c(3.1,3.1,0,0.1))
plot(1:200, CST, xlab = "", ylab = "$\\alpha_{\bot}^{'} z_t$", type = "l")
mtext(expression(italic(t)), side = 1, line=2)
```

#### Case 3: Intercept-only (Unrestricted Constant)
```{r}
nu0 <- matrix(c(0.5, -0.05), nrow = 2, ncol = 1)       # arbitrary numbers that don't happen to match those of the restricted intercept derived from mu0
for(i in 2:200){
   d.z_t[, i] <- nu0 + alpha %*% t(beta) %*% z_t[, i-1] + a_t[, i]
   z_t[, i]   <- z_t[, i-1] + d.z_t[, i]
}
```
#### Extracting the common trend and plotting
```{r}
# Extracting the common trend:
CST <- t(alpha.orth) %*% z_t 

# Plotting all into one picture:
par(mfrow=c(3, 1))
par(mar=c(0,3.1,3.1,0.1))
plot(z_t[1, ], type="l", xaxt="n", ylab = "$z_t$", main="", ylim = c(min(z_t), max(z_t)))
mtext("Unrestricted Constant", side = 3, line=1)
lines(z_t[2, ], col="red")
par(mar=c(0,3.1,0,0.1))
plot(1:200, t(beta) %*% z_t, type="l", xlab = "", ylab = "$\\beta^{'} z_t$", xaxt="n",)
par(mar=c(3.1,3.1,0,0.1))
plot(1:200, CST, xlab = "", ylab = "$\\alpha_{\bot}^{'} z_t$", type = "l")
mtext(expression(italic(t)), side = 1, line=2)
```

## Case 4: Intercept \& Trend (Restricted Trend)
```{r}
mu0 <- matrix(c(0.3, -0.2), nrow = 2, ncol = 1)      
mu1 <- matrix(c(0.3, 0.2), nrow = 2, ncol = 1)

nu  <- -alpha %*% t(beta) %*% mu0 + mu1
eta <- -t(beta) %*% mu1

for(i in 2:200){
   d.z_t[, i] <- nu + alpha %*% t(eta) * (i-1) +
       alpha %*% t(beta) %*% z_t[, i-1] + a_t[, i]
   z_t[, i]   <- z_t[, i-1] + d.z_t[, i]
}
```
#### Extracting the common trend and plotting
```{r}
# Extracting the common trend:
CST <- t(alpha.orth) %*% z_t 

# Plotting all into one picture:
par(mfrow=c(3, 1))
par(mar=c(0,3.1,3.1,0.1))
plot(z_t[1, ], type="l", xaxt="n", ylab = "$z_t$", main="", ylim = c(min(z_t), max(z_t)))
mtext("Restricted Trend", side = 3, line=1)
lines(z_t[2, ], col="red")
par(mar=c(0,3.1,0,0.1))
plot(1:200, t(beta) %*% z_t, type="l", xlab = "", ylab = "$\\beta^{'} z_t$", xaxt="n",)
par(mar=c(3.1,3.1,0,0.1))
plot(1:200, CST, xlab = "", ylab = "$\\alpha_{\bot}^{'} z_t$", type = "l")
mtext(expression(italic(t)), side = 1, line=2)
```

## Case 5: Intercept \& Trend (Unrestricted Intercept \& Trend)
```{r}
nu0  <- matrix(c(-0.1, -0.2), nrow = 2, ncol = 1)
nu1  <- matrix(c(0.001, 0.030), nrow = 2, ncol = 1)
print(cat("nu0:",nu0, "nu1:",nu1))

for(i in 2:200){
   d.z_t[, i] <- nu0 + nu1 * i + alpha %*% t(beta) %*% z_t[, i-1] + a_t[, i]
   z_t[, i]   <- z_t[, i-1] + d.z_t[, i]
}
```
#### Extracting the common trend and plotting
```{r}
# Extracting the common trend:
CST <- t(alpha.orth) %*% z_t 

# Plotting all into one picture:
par(mfrow=c(3, 1))
par(mar=c(0,3.1,3.1,0.1))
plot(z_t[1, ], type="l", xaxt="n", ylab = "$z_t$", main="", ylim = c(min(z_t), max(z_t)))
mtext("Unrestricted Trend", side = 3, line=1)
lines(z_t[2, ], col="red")
par(mar=c(0,3.1,0,0.1))
plot(1:200, t(beta) %*% z_t, type="l", xlab = "", ylab = "$\\beta^{'} z_t$", xaxt="n",)
par(mar=c(3.1,3.1,0,0.1))
plot(1:200, CST, xlab = "", ylab = "$\\alpha_{\bot}^{'} z_t$", type = "l")
mtext(expression(italic(t)), side = 1, line=2)
```




